Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-02-03 GECCO/volunteer-clouds-gecco16.tex
Minor fixes in experimental section

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of PUTs per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade or blue). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows an S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in Figure \ref{fig:puts:os}, which shows the
distribution of HTTP PUTs, equivalent to the number of
generations divided by 100, contributed by every user. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of PUTs per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of PUTs per worker and GEV fit (in light
  or blue color, left) 
and Weibull (in lighter or blue shade, right) for the {\sf NodIO-W$^2$} model.}  
\label{fig:gev:w2}
% File ../data/ips-puts.R
\end{figure}
%
As we did in the previous subsection, we have also fitted the number
of HTTP PUTs per worker to a  Generalized Extreme Value (GEV) function, with the result shown
in Table \ref{tab:puts:ww}. Comparing this table with what we obtained
in the previous experiment, Table \ref{tab:puts:os}, we find higher values
for the location $\mu$ and scale $\sigma$ parameters, accounting for
the higher number of contributions per volunteer that has been
previously observed. However, the shape $\xi$ parameter is
substantially similar, with a {\em bell} shape leaning towards the
origin, indicating a pattern of a few users with many contributions
and many with few contributions. If we plot the model and the data
side by side, as shown in Figure \ref{fig:gev:w2}, we see that the fit
is not so tight as in the previous case, with the model overestimating
the number of individuals with a high number contributions. This is
probably due to the fact that in the previous case the user needed to
voluntarily reload the page to contribute the most generations, with a
{\em phase change} between the people that did so and the people that
just ran the experiment once until completion. In this case, however, the
user just needs to let it run, without that phase change and thus we obtain a
more {\em egalitarian} distribution of contributions, which rather
corresponds to a Weibull function, as shown in Figure
\ref{fig:ipstime:w2} (right hand side). The fitted parameters for this
distribution, shown in Table \ref{tab:puts:ww}, show a shape parameter
equal to 0.69, a value that is remarkably similar to the 0.5 value
found for time devoted to games in Chambers et
al. \cite{chambers2005measurement}. It is important to note that, in
this case, the number of PUTs made per client does not correspond to a
uniform amount of computation, since 100 generations with a random
population belong, in each case, to different number of
operations. In other words, the same number of PUTs will correspond to
different number of operations in each case. However, since clients
are also different and take a different time in each case, it is
difficult to ascertain how this might have an influence in the
statistical distribution.

@@

The fact that the solution of this particular problem was found in a
few seconds and restarts were necessary yielded interesting results, % (Paloma) Where?
but we needed to work on a harder problem, that is why created a new
client version that optimized a harder variation of the Rastrigin's function, 
used as a benchmark in evolutionary
algorithms. The set up for this function, and a comparison of the
baseline speed for JavaScript in an standalone experiment and other
languages is shown in \cite{2016arXiv160101607Manom}, where the details of
the implementation have also been extensively described.

Specifically, this experiment was deployed in a virtual private server with 512MB of memory
running Ubuntu OS version 14.04 and hosted in DigitalOcean.com. Unlike previous
experiments where the optimum was expected to be found many times, 
this problem was expected to constantly run for many days finding only
sub-optimal approximations. In order to achieve this, the only change needed 
in the framework was to limit the number of chromosomes kept in the pool. 
Previously, every PUT request which included a new chromosome increased 
the size of the pool. But in a long-running experiment this has to be avoided
in order to not exhausting the available memory. In this experiment the size
of a chromosome was 1000 double precision numbers, so that considering this
and the memory available in the server, the pool was limited 
to 10,000 chromosomes.  In order to keep this size
constant, a random chromosome was pulled out of the pool before inserting a
new one, when the pool was full. For this problem {\sf NodEO}'s {\tt chromosome-float} 
and {\tt classic-ea-float} modules were used. The parameters for the EA 
were tournament size = 3 and  population size = 500, returning exchanging a
chromosome with the server every 100 generations. Again, there were two workers in each window.
As in previous experiments, a few calls to participation were published in social networks.

@@

The results for this new experiment have been extremely interesting
and quite different from the others. The most important
characteristic that set this experiment apart is
the fact that it did not find the solution for the time it was
running, instead lowering, little by little, the fitness until it was
disconnected. The fitness of chromosomes sent by clients managed to
surpass the 9000 value by the end of the execution; and although initially there
was a high number of new nodes, this number decreased with time.

>>>>

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of PUTs per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade or blue). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows an S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, on the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in Figure \ref{fig:puts:os}, which shows the
distribution of HTTP PUTs, equivalent to the number of
generations divided by 100, contributed by every user. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of PUTs per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
%Pablo: question for Paloma: is over all or overall?
into account the low correlation between successive events shown
above.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of PUTs per worker and GEV fit (in light
  or blue color, left) 
and Weibull (in lighter or blue shade, right) for the {\sf NodIO-W$^2$} model.}  
\label{fig:gev:w2}
% File ../data/ips-puts.R
\end{figure}
%
As we did in the previous subsection, we have also fitted the number
of HTTP PUTs per worker to a  Generalized Extreme Value (GEV) function, with the result shown
in Table \ref{tab:puts:ww}. Comparing this table with what we obtained
in the previous experiment, Table \ref{tab:puts:os}, we find higher values
for the location $\mu$ and scale $\sigma$ parameters, accounting for
the higher number of contributions per volunteer that has been
previously observed. However, the shape $\xi$ parameter is
substantially similar, with a {\em bell} shape leaning towards the
origin, indicating a pattern of a few users with many contributions
and many with few contributions. If we plot the model and the data
side by side, as shown in Figure \ref{fig:gev:w2}, we see that the fit
is not so tight as in the previous case, with the model overestimating
the number of individuals with a high number contributions. This is
probably due to the fact that in the previous case the user needed to
voluntarily reload the page to contribute the most generations, with a
{\em phase change} between the people that did so and the people that
just ran the experiment once until completion. In this case, however, the
user just needs to let it run, without that phase change and thus we obtain a
more {\em egalitarian} distribution of contributions, which rather
corresponds to a Weibull function, as shown in Figure
\ref{fig:ipstime:w2} (right hand side). The fitted parameters for this
distribution, shown in Table \ref{tab:puts:ww}, show a shape parameter
equal to 0.69, a value that is remarkably similar to the 0.5 value
found for time devoted to games in Chambers et
al. \cite{chambers2005measurement}. It is important to note that, in
this case, the number of PUTs made per client does not correspond to a
uniform amount of computation, since 100 generations with a random
population belong, in each case, to different number of
operations. In other words, the same number of PUTs will correspond to
different number of operations in each case. However, since clients
are also different and take a different time in each case, it is
difficult to ascertain how this might have an influence on the
statistical distribution.

@@

The fact that the solution of this particular problem was found in a
few seconds and restarts were necessary yielded interesting results, % (Paloma) Where?
but we needed to work on a harder problem, that is why created a new
client version that optimized a harder variation of the Rastrigin's function, 
used as a benchmark in evolutionary
algorithms. The set up for this function, and a comparison of the
baseline speed for JavaScript in a standalone experiment and other
languages is shown in \cite{2016arXiv160101607Manom}, where the details of
the implementation have also been extensively described.

Specifically, this experiment was deployed in a virtual private server with 512MB of memory
running Ubuntu OS version 14.04 and hosted in DigitalOcean.com. Unlike previous
experiments where the optimum was expected to be found many times, 
this problem was expected to constantly run for many days finding only
sub-optimal approximations. In order to achieve this, the only change needed 
in the framework was to limit the number of chromosomes kept in the pool. 
Previously, every PUT request which included a new chromosome increased 
the size of the pool. But in a long-running experiment this has to be avoided
in order to not exhaust the available memory. In this experiment the size
of a chromosome was 1000 double precision numbers, so that considering this
and the memory available in the server, the pool was limited 
to 10,000 chromosomes.  In order to keep this size
constant, a random chromosome was pulled out of the pool before inserting a
new one, when the pool was full. For this problem {\sf NodEO}'s {\tt chromosome-float} 
and {\tt classic-ea-float} modules were used. The parameters for the EA 
were tournament size = 3 and  population size = 500, returning exchanging a
chromosome with the server every 100 generations. Again, there were two workers in each window.
As in previous experiments, a few calls to participation were published on social networks.

@@

The results for this new experiment have been extremely interesting
and quite different from the others. The most important
characteristic that sets this experiment apart is
the fact that it did not find the solution for the time it was
running, instead lowering, little by little, the fitness until it was
disconnected. The fitness of chromosomes sent by clients managed to
surpass the 9000 value by the end of the execution; and although initially there
was a high number of new nodes, this number decreased with time.
